Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography

### 2023-08-24 manuscript/wp4.tex
ms aug 24

\title{How does the writing system get informative? Differentiation or conservation?}

\shorttitle{Informativity in the writing system}

@@

\abstract{Abstract.}

@@

Learnability, measured as transmission error (lower error = higher learnability). Mean Levenshtein edit distance between the words produced at generation i and the corresponding words produced at generation i-1.

Simplicity, measured as complexity (lower complexity = higher simplicity). The length (in bits) of the shortest grammar that losslessly compresses the lexicon, as computed by the Grammarette package: https://github.com/jwcarr/grammarette. Finding the shortest grammar for some lexicon is a non-trivial problem, and it is likely that we will need to adapt the method in response to the collected data.

Communicative cost is given by
\begin{equation}
\mathrm{cost}(L) := \frac{1}{|U|} \sum_{m \in U} -\log \mathrm{Pr}(m|s_m),
\end{equation}
where $\mathrm{Pr}(m|s_m)$ is the probability that a listener would infer meaning $m$ given that a speaker produced signal $s$ for meaning $m$. Here we assume this probability to be $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$.

To analyze the variation in the suffix spellings and how it correlates with meaning, each participant's system was transformed into a 3×3 matrix, with rows representing the shape dimension, columns representing the color dimension, and the matrix values representing the unique spellings. For example, a ``degenerate'' system of suffix spellings---one that uses a single spelling for all meanings---would have the following matrix representation:
\begin{equation}
	S_\mathrm{degenerate} = \begin{bmatrix}
	1 & 1 & 1 \\
	1 & 1 & 1 \\
	1 & 1 & 1
	\end{bmatrix}
\end{equation}
Here the number 1 represents the one and only suffix spelling. We define an ``expressive'' system as one that marks the color dimension using three distinct spellings, resulting in the following matrix:
\begin{equation}
	S_\mathrm{expressive} = \begin{bmatrix}
	1 & 2 & 3 \\
	1 & 2 & 3 \\
	1 & 2 & 3
	\end{bmatrix}
\end{equation}
In such a system, although the suffix only expresses color, the words as a whole are fully informative, since shape is expressed by the stem. A third system of interest is the ``redundant'' system:
\begin{equation}
	S_\mathrm{redundant} = \begin{bmatrix}
	1 & 1 & 1 \\
	2 & 2 & 2 \\
	3 & 3 & 3
	\end{bmatrix}
\end{equation}
Although $S_\mathrm{redundant}$ carries the same amount of information as $S_\mathrm{expressive}$, it marks a dimension (shape) that is already expressed by the stem, making the information carried by the suffix redundant. Such a system contains unnecessary complexity---different ways of writing /kəʊ/ that carry no additional information.

Of course, the suffix spelling systems produced by participants may not match these three reference systems exactly. To determine which reference system a participant's output system is most similar to, we adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric. VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another. So, for example, the distance between $S_\mathrm{degenerate}$ and $S_\mathrm{expressive}$ is 2~bits because $S_\mathrm{expressive}$ conveys 2~bits of information\footnote{In $S_\mathrm{expressive}$, knowing the suffix reduces uncertainty from 16 equally probable meanings to 4 equally probable meanings; $-\log\frac{4}{16} = 2$~bits.}, 2~bits more than $S_\mathrm{degenerate}$, which conveys 0~bits of information. The distance between $S_\mathrm{expressive}$ and $S_\mathrm{redundant}$ is 4~bits; this is because, although both systems convey 2~bits of information, the information they convey is entirely orthogonal; in switching from $S_\mathrm{expressive}$ to $S_\mathrm{redundant}$, one loses 2~bits of information (information about color), but one gains 2~bits of information (information about shape), making the total distance 4~bits.

We computed the VoI between the emergent suffix systems and the three reference systems and plotted them on a ternary diagram, the three vertices of which represent the three reference systems (see Fig.~\ref{exp1_ternary_plot}). The diagram is divided into three sectors (shades of gray), which reveal which reference system the final converged-on suffix systems are most similar to. Most of the emergent system are either redundant or degenerate, with some straddling the border between the two. No system is classified as expressive, although two of the converged-on systems (highlighted in Fig.~\ref{exp1_ternary_plot}) showed some signs of color-marking. In one system, <icoe> is used for gray objects and <ikoe> for all other colors. In the other system, <ikcow> is used for gray objects and <ikoe> is used for all other colors, except the star shape uses <y> instead of <i> (i.e., <ykcow> and <ykoe>), resulting in some spelling redundancy\footnote{However, rather than parsing the stems as <buv>, <zett>, <gaf>, <wop>, they could instead be parsed as <buvi>, <zetty>, <gafi>, <wopi>. Under this interpretation, there would be no redundancy and the system would be equivalent to the other 1-color emergent system.} In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the Transmission-only condition.

@@

Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation. Exposed to an orthography with high variation, learners could have gradually conditioned that variation on the dimension where it would have been useful from the point of view of disambiguation, color, yielding differentiated forms in the orthography. But this did not happen, even under communicative pressure. Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure. So, what was different about the present experiment? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative, which acts as a cue to participants that the language does not mark color and that, therefore, the orthography should not mark color. Faced with variation that could be conditioned on shape or color, learners rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of the /ʃən/ suffix (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion}, \textit{transmission}, and \textit{invention}).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/con_com.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission~+ Communication condition in Experiment~2 (conservation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items.}
	\label{con_com}
	\end{figure}

@@

@@

In contrast with Transmission-only, where only one chain remained expressive into Epoch~III, in Transmission~+ Communication, five chains (O,~P, Q, R, and~T) remained expressive, six if you include Chain~B, whose <sol> spelling does find its origin in the <soe> spelling from the seed orthography.

>>>>

\title{Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography}

\shorttitle{The cultural evolution of informative orthography}

@@

\abstract{It is widely acknowledged that opaque orthographies place additional demands on learning, often requiring many years to acquire fully. It is less widely recognized, however, that such orthographies may offer certain benefits in the context of reading. For example, the heterographic homophones (e.g., <knight> vs. <night>) impose additional costs in learning but reduce ambiguity in reading. Could it be the case, then, that heterography---and therefore orthographic opacity---may sometimes be selected for in the evolution of writing systems, despite the cost in learning? We investigate this question by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for avoiding ambiguity. In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate words that are homophonous in speech (e.g., <lite> from <light> to signal low-fat), and \textit{conservation}, in which heterography arises through exaptation as an epiphenomenon of sound change (e.g., <meat> vs. <meet> resulting from the /ɛː/–/eː/ merger in the Great Vowel Shift). Under pressure from learning alone, orthographic systems become transparent, but when combined with communicative pressure, these systems can become informative, diverging from the spoken language in order to express meaning directly.}

@@

As a quantitative measure of how informative the orthographic systems are over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}. Communicative cost expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. Formally, it is given by $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$,
	% \begin{equation}
	% \sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
	% \end{equation}
where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.

Fig.~\ref{cost} plots communicative cost over the nine generations of cultural transmission for both the Transmission-only condition (left) and Transmission + Communication condition (right). Each colored line represents an individual chain, and the thick black lines shows the mean across chains. The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling. In both conditions, cost increases with generation, implying that the languages become less informative over time. This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent. In contrast, this is not the expected result in Transmission + Communication, where the communicative pressure should promote the emergence of informative systems, but it does align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produce differentiated spellings, be it through explicit innovation or implicit generalization. Note, however, that the cost plateaued at around 1.3~bits in Transmission-only and around 1~bit in Transmission + Communication. This suggests that there is some tendency for the systems to remain more informative in the communicative condition.

To evaluate this argument statistically, we regressed cost on generation in a Bayesian hierarchical cubic regression model with a gamma likelihood and log link function. The posterior parameter estimates are shown in Fig.~\ref{stats1}. Essentially, the beta coefficients of the model (representing the shape of the polynomial) do not differ between conditions, but the intercepts do differ (generation was centered, such that the intercept corresponds to generation 5). This difference in intercepts reflects the fact that the cost plateaued at a lower level in the Transmission + Communication condition. Nevertheless, the posteriors for $\alpha_\mathrm{tran}$ and $\alpha_\mathrm{comm}$ do overlap to some extent, suggesting no definitive difference between conditions.

To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate, as illustrated in Fig.~\ref{}. These four typological categories form a continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning. The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing. To classify a participant's output into one of these four categories, we may compute which of the reference systems is most similar in terms of its information content. To do this, we formalize the systems as set partitions (i.e., each system may be viewed as a partitioning of the universe of meanings into $n$ disjoint sets) and adopt variation of information \parencite[VoI;][]{Meila:2007} as a distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another.}

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_dif.eps}}
	\vspace*{2pt}
	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow). At generation~0, all 10 chains in both conditions use a holistic system (by design). In Transmission-only, the systems switch first to a redundant and then to a degenerate patterning. In Transmission + Communication, the systems ultimately become degenerate, but also tend to favor holistic systems over redundant ones along the way. Expressive systems were rare in both conditions.}
	\label{typ_dist_dif}
	\end{figure}

Rather than simply classify each system into exactly one typological category, we computed probability distributions over the typological space according to distance. This gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially redolent of one category but also redolent of another (for example, system T9 is part redundant and part degenerate). The probability distribution was calculated as $e^{-3\cdot\mathrm{VOI}(p, q)}$

Fig.~\ref{typ_dist_dif} plots the typological distribution (i.e., across chains) for each generation and each condition.

We computed the VoI between the emergent suffix systems and the three reference systems and plotted them on a ternary diagram, the three vertices of which represent the three reference systems (see Fig.~\ref{exp1_ternary_plot}). The diagram is divided into three sectors (shades of gray), which reveal which reference system the final converged-on suffix systems are most similar to. Most of the emergent system are either redundant or degenerate, with some straddling the border between the two. No system is classified as expressive, although two of the converged-on systems (highlighted in Fig.~\ref{exp1_ternary_plot}) showed some signs of color-marking. In general, there is no discernible difference between the two conditions; regardless of whether the communicative pressure was present, the emergent suffix spelling systems tended to fall along the degenerate--redundant axis, and the only two cases of color-marking were in~fact from the Transmission-only condition.

@@

Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation. Exposed to an orthography with high variation, learners could have gradually conditioned that variation on the dimension where it would have been useful from the point of view of disambiguation, color, yielding differentiated forms in the orthography. But this did not happen, even under communicative pressure. Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure. So, what was different about the present experiment? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative, which acts as a cue to participants that the language does not mark color and that, therefore, the orthography should not mark color. Faced with variation that could be conditioned on shape or color, learners rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion}, \textit{transmission}, and \textit{invention}).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/con_com.eps}}
	\vspace*{2pt}
	\caption{Results from the Transmission~+ Communication condition in Experiment~2 (conservation). Each matrix shows the suffix spelling system in use at a particular generation (shape on the rows, color on the columns, as in Fig.~\ref{stimuli}). Chains are labeled K--T and generations are labeled 0–9 (0 is the randomly generated seed system). Each chain uses an independent color palette, with each color representing a particular suffix spelling; similar colors indicate similar spellings. Spellings in bold-italic are the generalizations on unseen items. Five chains (O, P, Q, R, T) remain fully expressive into the final epoch, in most cases conserving the original forms.}
	\label{con_com}
	\end{figure}

@@

As in Experiment~1, we plot communicative cost by generation in both conditions (see Fig.~\ref{cost2}). Each colored line represents an individual chain, and the thick black lines shows the mean across chains. These mean lines reveal a fairly clear effect from epoch. During Epoch~I, there is a shallow increase in cost with generation in both conditions. This represents a kind of baseline level of noise when the spoken languages are fully informative. During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in Transmission + Communication. This is because the communicative pressure acts to slow down the rate at which the systems become transparent or redundant. In Epoch~III, the cost curve plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change. In Transmission + Communication, by contrast, the curve gets a little steeper because the systems still have some room to simplify and the pressure to simplify has been ramped up.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_con.eps}}
	\vspace*{2pt}
	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow). At generation~0, all 10 chains in both conditions use an expressive system (by design). In Transmission-only, the systems tend to transition to the redundant or degenerate patterning, while in Transmission + Communication, they tend to remain expressive for longer and disfavor the redundant pattern.}
	\label{typ_dist_con}
	\end{figure}

@@

In contrast with Transmission-only, where only one chain remained expressive into Epoch~III, in Transmission~+ Communication, five chains (O,~P, Q, R, and~T) remained expressive, or six if you include Chain~B, whose <sol> spelling does find its origin in the <soe> spelling from the seed orthography.
